
# Life-history traits explain variation in the larval dispersal potential of corals - systematic review figure code


## Meta-analysis models and figures


Code to re-produce our meta-analysis and to produce figure 4



### Packages

Code to load (and install if necessary) the packages needed. (Remove "#"s to install packages.)
```{r, message=FALSE}
#  install.packages("tidyverse")
library(tidyverse)

# install.packages("metafor")
library(metafor)

# install.packages("rotl")
library(rotl)

# install.packages("ape")
library(ape)
```



### Data


Read in the effect size data
```{r, message=FALSE}
#Review data
model_data_raw <- read_csv("data/coral-modelling-data.csv")
```


Separate the mean settlement and survival times effect sizes into separate dataframes
```{r}
coral_settle <- model_data_raw %>% 
  drop_na(pld_av_comp_effectsize_d)

coral_survival <- model_data_raw %>% 
  drop_na(pld_av_mort_effectsize_d)
```



### Data wrangling - computing phylogeny


Make a phylogenetic correlation matrix for the species with competency data (can be slow!)
```{r}
#Create list of species for meta-analysis
settle_splist <- coral_settle %>% 
  distinct(species_accepted, .keep_all = FALSE)
#Make list a vector
settle_splist <- settle_splist$species_accepted

#Get ottd ids for species (this can take a minute)
ott_ids_settle <- tnrs_match_names(settle_splist) #%>% 
  #drop_na(ott_id)

print(ott_ids_settle)
#They all match, yay

#Make tree
settle_tree <- tol_induced_subtree(ott_ids_settle$ott_id)
plot(settle_tree)

#Make tree ultrametric
settle_tree <- compute.brlen(settle_tree)

#Make a phylogenetic correlation matrix from the tree
settle_PCM <- vcv(settle_tree, corr = TRUE)

```

Save a plot of the tree
```{r}
plot(settle_tree)
png(filename = "figures/settlement_tree.png",
    width = 900, height = 1500)
plot(settle_tree)
dev.off()
```


Match the ottd IDs to the settlement data
```{r}
#Add the species OTTD IDs to link to the phylogenetic correlation matrix
settle_link <- subset(ott_ids_settle, select = c("search_string", "unique_name", "ott_id"))
phy_settle <- coral_settle %>% 
  mutate(low_species = tolower(species_accepted)) %>% 
  mutate(low_genus = tolower(genus_accepted))
phy_settle <- left_join(x = phy_settle, y = settle_link, by = join_by(low_species == search_string))

#change the unique name for A. millepora to put underscores in the bracketed bit
phy_settle <- phy_settle %>% 
  mutate(unique_name = case_when(
    grepl("millepora", unique_name) ~ "Acropora_millepora_(species_in_domain_Eukaryota)",
    .default = unique_name
  ))

phy_settle <- phy_settle %>% 
  mutate(species.phy = sub(" ", "_", (paste(unique_name, paste("ott", phy_settle$ott_id, sep = ""), sep = "_"))))

```



Now do the same for the survival data
```{r}
#Create list of species for meta-analysis
surv_splist <- coral_survival %>% 
  distinct(species_accepted, .keep_all = FALSE)
#Make list a vector
surv_splist <- surv_splist$species_accepted

#Get ottd ids for species (this can take a minute)
ott_ids_surv <- tnrs_match_names(surv_splist) #%>% 
  #drop_na(ott_id)

print(ott_ids_surv)
#They all match, yay

#Make tree
surv_tree <- tol_induced_subtree(ott_ids_surv$ott_id)
plot(surv_tree)

#Make tree ultrametric
surv_tree <- compute.brlen(surv_tree)

#Make a phylogenetic correlation matrix from the tree
surv_PCM <- vcv(surv_tree, corr = TRUE)

```


Save a plot of the tree
```{r}
plot(surv_tree)
png(filename = "figures/survival_tree.png",
    width = 900, height = 900)
plot(surv_tree)
dev.off()
```


Match the ottd IDs to the survival data
```{r}
#Add the species OTTD IDs to link to the phylogenetic correlation matrix
surv_link <- subset(ott_ids_surv, select = c("search_string", "unique_name", "ott_id"))
phy_surv <- coral_survival %>% 
  mutate(low_species = tolower(species_accepted)) %>% 
  mutate(low_genus = tolower(genus_accepted))
phy_surv <- left_join(x = phy_surv, y = surv_link, by = join_by(low_species == search_string))

#change the unique name for A. millepora to put underscores in the bracketed bit
phy_surv <- phy_surv %>% 
  mutate(unique_name = case_when(
    grepl("millepora", unique_name) ~ "Acropora_millepora_(species_in_domain_Eukaryota)",
    .default = unique_name
  ))

phy_surv <- phy_surv %>% 
  mutate(species.phy = sub(" ", "_", (paste(unique_name, paste("ott", phy_surv$ott_id, sep = ""), sep = "_"))))

```



Make reproductive mode and morphology factors so the models work properly
```{r}
#Reproductive mode
phy_settle$spawn_mode <- as.factor(phy_settle$spawn_mode)
phy_surv$spawn_mode <- as.factor(phy_surv$spawn_mode)

#Morphology
phy_settle$growth_form_cat <- as.factor(phy_settle$growth_form_cat)
phy_surv$growth_form_cat <- as.factor(phy_surv$growth_form_cat)
```


Create dataframes with geographic variables for models of geographic realm and latitude, for both settlement and survival time effect sizes
```{r}
# Settlement
geog_settle <- phy_settle %>% 
  mutate(zone = case_when(
    grepl("Temperate", biogeog_realm) ~ "temperate",
    grepl("Tropical", biogeog_realm) ~ "tropical",
    grepl("Indo-Pacific", biogeog_realm) ~ "tropical",
    grepl("Arctic", biogeog_realm) ~ "polar",
    grepl("Southern Ocean", biogeog_realm) ~ "polar")) %>% # assign realms to tropical/temperate
  mutate(med_lat = (abs(latmin) + abs(latmax))/2) %>% # determine latitude of the study area's midpoint
  filter(biogeog_realm != "aquarium_culture") # remove effect sizes for observations with unknown locations

# Survival
geog_surv <- phy_surv %>% 
  mutate(zone = case_when(
    grepl("Temperate", biogeog_realm) ~ "temperate",
    grepl("Tropical", biogeog_realm) ~ "tropical",
    grepl("Indo-Pacific", biogeog_realm) ~ "tropical",
    grepl("Arctic", biogeog_realm) ~ "polar",
    grepl("Southern Ocean", biogeog_realm) ~ "polar",
  )) %>% 
  mutate(med_lat = (abs(latmin) + abs(latmax))/2) %>% 
  filter(biogeog_realm != "aquarium_culture")
```




### Random-effects meta-analysis models: Settlement time


Run random-effects meta-analysis models of mean larval settlement time.


Random-effects model - observation nested in study
```{r}
# Construct model
random_study_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = phy_settle, random = list(~ 1|acc_id/obs_id)) # use mean settlement time as effect size, weight by variance in the mean, use observation as a random effect nested within study

# Summarise results
summary(random_study_mod_SETTLE)
```


Random-effects model - observation nested in study with phylogenetic control
```{r}
# Model
random_phylo_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = phy_settle, random = list(~ 1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = settle_PCM)) # add species as a random effect, and link it to the relevant phylogenetic correlation matrix

# Results
summary(random_phylo_mod_SETTLE)
```


AIC comparison shows that controlling for phylogeny provides better fit



### Mixed-effects meta-regressions: Settlement time


Combined reproductive mode and morphology model
```{r}
repro_morpho_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = phy_settle, mods = ~ spawn_mode*growth_form_cat, random = list(~ 1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = settle_PCM))

summary(repro_morpho_mod_SETTLE)
```



Model for biogeographic realm
```{r}
realm_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = geog_settle, mods = ~biogeog_realm, random = list(~1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = settle_PCM))

summary(realm_mod_SETTLE)
```


Model for tropical/temperate region
```{r}
geog_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = geog_settle, mods = ~zone, random = list(~1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = settle_PCM))
summary(geog_mod_SETTLE)
```


Model for latitude
```{r}
lat_mod_SETTLE <- rma.mv(yi = pld_av_comp_effectsize_d, V = ((pld_av_comp_effect_sd)^2), data = geog_settle, mods = ~med_lat, random = list(~1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = settle_PCM))
summary(lat_mod_SETTLE)

```



### Random-effects meta-analysis models: Survival time


Run random-effects meta-analysis models of mean larval settlement time.


Random-effects model - observation nested in study
```{r}
random_study_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = phy_surv, random = list(~ 1|acc_id/obs_id))

summary(random_study_mod_SURV)
```


Random-effects model - observation nested in study with phylogenetic control
```{r}
random_phylo_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = phy_surv, random = list(~ 1|acc_id/obs_id, ~1|species.phy), R = list(species.phy = surv_PCM))

summary(random_phylo_mod_SURV)
```


AIC comparison shows that controlling for phylogeny does not provide better fit



### Mixed-effects meta-regressions: Survival time


Combined reproductive mode and morphology model
```{r}
repro_morpho_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = phy_surv, mods = ~ spawn_mode*growth_form_cat, random = list(~ 1|acc_id/obs_id))

summary(repro_morpho_mod_SURV)
```



Model for biogeographic realm
```{r}
realm_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = geog_surv, mods = ~biogeog_realm, random = list(~1|acc_id/obs_id))

summary(realm_mod_SURV)
```


Model for tropical/temperate region
```{r}
geog_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = geog_surv, mods = ~zone, random = list(~1|acc_id/obs_id))
summary(geog_mod_SURV)
```


Model for latitude
```{r}
lat_mod_SURV <- rma.mv(yi = pld_av_mort_effectsize_d, V = ((pld_av_mort_effect_sd)^2), data = geog_surv, mods = ~med_lat, random = list(~1|acc_id/obs_id))
summary(lat_mod_SURV)
```




### Data wrangling - figure 4


Produce a dataframe with mean settlement time estimates for each reproductive mode and morphology combination as predicted by the model, with 95% confidence intervals
```{r}
colnames(repro_morpho_mod_SETTLE$X)

grid <- expand.grid(
  spawn_mode = levels(phy_settle$spawn_mode),
  growth_form_cat = levels(phy_settle$growth_form_cat)
)

mm <- model.matrix(~ spawn_mode * growth_form_cat, data = grid)

mm <- mm[, -1, drop = FALSE]

# match model columns
keep <- colnames(repro_morpho_mod_SETTLE$X)[-1]  # remove intrcpt
mm <- mm[, keep, drop = FALSE]

preds <- predict(repro_morpho_mod_SETTLE, newmods = mm)

result <- cbind(grid, preds)
result
```



Re-order and relabel settlement time data for plotting. Add in upper and lower confidence intervals for each data point, using the SE reported
```{r}
settlement_plotting <- phy_settle %>% 
  mutate(morph_id = case_when(
    growth_form_cat == "branching" ~ "1",
    growth_form_cat == "domed" ~ "2",
    growth_form_cat == "plating" ~ "3",
    growth_form_cat == "erect branched" ~ "4",
    growth_form_cat == "horizontal unbranched" ~ "5"
    )) %>% 
  mutate_at("morph_id", as.numeric) %>% # Assign each morphology a number for re-ordering them
  mutate(spawn_mode = factor(spawn_mode, 
         levels = c("broadcasting", "brooding"), 
         labels = c("Broadcasting", "Brooding"))) %>% # Re-label reproductive mode
  mutate(growth_form_cat = factor(growth_form_cat, 
         levels = c("branching", "domed", "plating", "erect branched", "horizontal unbranched"), 
         labels = c("Branching (stony)", "Domed (stony)", "Plating (stony)", "Erect branched (soft)", "Horizontal unbranched (soft)"))) %>% # Re-label morphology
  mutate_at("species_accepted", as.factor) %>%
  mutate(comp_lb = pld_av_comp_effectsize_d + 1.96*pld_av_comp_effect_sd) %>% 
  mutate(comp_ub = pld_av_comp_effectsize_d - 1.96*pld_av_comp_effect_sd) # Add confidence intervals
```


Do the same for survival time data
```{r}
survival_plotting <- phy_surv %>% 
  mutate(morph_id = case_when(
    growth_form_cat == "branching" ~ "1",
    growth_form_cat == "domed" ~ "2",
    growth_form_cat == "plating" ~ "3",
    growth_form_cat == "erect branched" ~ "4",
    growth_form_cat == "horizontal unbranched" ~ "5"
  )) %>% 
  mutate_at("morph_id", as.numeric) %>% 
  mutate(growth_form_cat = factor(growth_form_cat, 
         levels = c("branching", "domed", "plating", "erect branched", "horizontal unbranched"), 
         labels = c("Branching (stony)", "Domed (stony)", "Plating (stony)", "Erect branched (soft)", "Horizontal unbranched (soft)"))) %>% 
  mutate(spawn_mode = factor(spawn_mode, 
         levels = c("broadcasting", "brooding"), 
         labels = c("Broadcasting", "Brooding"))) %>%
  mutate_at("species_accepted", as.factor) %>%  
  mutate(surv_ub = pld_av_mort_effectsize_d + 1.96*pld_av_mort_effect_sd) %>%    
  mutate(surv_lb = pld_av_mort_effectsize_d - 1.96*pld_av_mort_effect_sd) # Add confidence intervals
```



Manually set shapes for points for each reproductive mode
```{r}
#Settlement time effect sizes
repro_shape_set <- c(1, 19)
names(repro_shape_set) <- levels(settlement_plotting$spawn_mode)

#Survival time effect sizes
repro_shape_surv <- c(1, 19)
names(repro_shape_surv) <- levels(survival_plotting$spawn_mode)
```



### Figure 4


Produce 4a, a forest plot showing variation in mean larval settlement time effect sizes, with vertical lines showing overall mean settlement times for each life-history group, as estimated by the meta-regression.
```{r}
# Create plot
fig4a <- ggplot(settlement_plotting, aes(x = pld_av_comp_effectsize_d, y = fct_reorder(species_accepted, morph_id, .desc = TRUE), shape = spawn_mode, colour = growth_form_cat)) + # Plot species, grouped by morphology, against mean larval settlement time, with shape and colour differentiating reproductive mode and morphology
  geom_errorbar(aes(xmin = comp_lb, xmax = comp_ub), alpha = 0.55, width = 0.5) + # Add error bars for data points
  geom_point(alpha = 0.7, size = 3) + # Set point size and transparency
  geom_hline(yintercept = 2.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 14.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 21.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 52.5, colour = "grey", linetype = 2, linewidth = 1.5) + # Use dashed lines to separate out morphologies
  geom_segment(aes(x = 5.264, y = 77.5, yend = 52.5), colour = "#5C4564", linewidth = 1) + # Add vertical lines for overall means for each life-history grouping, and show confidence interval
  annotate(geom = "rect", xmin = 0.001, xmax = 10.526	,
           ymin = 77.5, ymax = 52.5,
           fill = "#5C4564", alpha = 0.2) +
  geom_segment(aes(x = 13.873, y = 77.5, yend = 52.5), colour = "#5C4564", linetype = 2, linewidth = 1) +
  annotate(geom = "rect", xmin = 10.043, xmax = 17.703,
           ymin = 77.5, ymax = 52.5,
           fill = "#5C4564", alpha = 0.2) +
  geom_segment(aes(x = 8.740, y = 52.5, yend = 21.5), colour = "#41C3EE", linewidth = 1) +
  annotate(geom = "rect", xmin = 1.838, xmax = 15.642,
           ymin = 21.5, ymax = 52.5,
           fill = "#41C3EE", alpha = 0.15) +
  geom_segment(aes(x = 12.988, y = 52.5, yend = 21.5), colour = "#41C3EE", linetype = 2, linewidth = 1) +
  annotate(geom = "rect", xmin = 8.889, xmax = 17.088,
           ymin = 21.5, ymax = 52.5,
           fill = "#41C3EE", alpha = 0.15) +
  geom_segment(aes(x = 14.093, y = 14.5, yend = 21.5), colour = "#AFFC57", linetype = 2, linewidth = 1) +
  annotate(geom = "rect", xmin = 6.865, xmax = 21.321,
           ymin = 14.5, ymax = 21.5,
           fill = "#AFFC57", alpha = 0.1) +
  geom_segment(aes(x = 29.112, y = 2.5, yend = 14.5), colour = "#FB9140", linewidth = 1) +
  annotate(geom = "rect", xmin = 17.870, xmax = 37.574,
           ymin = 2.5, ymax = 14.5,
           fill = "#FB9140", alpha = 0.15) +
  geom_segment(aes(x = 27.722, y = 2.5, yend = 14.5), colour = "#FB9140", linetype = 2, linewidth = 1) +
  annotate(geom = "rect", xmin = 22.523, xmax = 35.701,
           ymin = 2.5, ymax = 14.5,
           fill = "#FB9140", alpha = 0.15) +
  geom_segment(aes(x = 9.922, y = 0.5, yend = 2.5), colour = "#973939", linewidth = 1) +
  annotate(geom = "rect", xmin = 0, xmax = 20.537,
           ymin = 0.5, ymax = 2.5,
           fill = "#973939", alpha = 0.15) +
  scale_shape_manual(values = repro_shape_set,
                     name = "Reproductive mode") + # Set point shapes manually
  scale_colour_viridis_d(option = "H", name = "Colony morphology") + # Use an accessible colour pallette
  theme_bw() + # Set theme manually
  guides(
    shape = guide_legend(order = 1),
    colour = guide_legend(order = 2)
  ) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(vjust = 0.5),
        axis.title.x = element_text(vjust = -0.5),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        plot.title = element_text(size = 14, colour = "black", face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black", face = "bold"),
        legend.position = "inside",
        legend.position.inside = c(0.8, 0.814)) +
  labs(x = "Mean larval settlement time / days (95% confidence interval)",
       y = "Species")  # Label axes

#View figure
fig4a

#Save figure
ggsave("figures/fig4a.png", width = 12, height = 17)
```



Make figure for survival times (4b). No vertical lines needed for averages.
```{r}
# Create figure
fig4b <- ggplot(survival_plotting, aes(x = pld_av_mort_effectsize_d, y = fct_reorder(species_accepted, morph_id, .desc = TRUE), shape = spawn_mode, colour = growth_form_cat)) +
  geom_errorbar(aes(xmin = surv_lb, xmax = surv_ub), alpha = 0.55, width = 0.5) +
  geom_point(alpha = 0.7, size = 3) + 
  geom_hline(yintercept = 1.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 8.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 12.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  geom_hline(yintercept = 24.5, colour = "grey", linetype = 2, linewidth = 1.5) + 
  scale_shape_manual(values = repro_shape_surv,
                     name = "Reproductive mode") + 
  scale_colour_viridis_d(option = "H", name = "Colony morphology") +
  theme_bw() +
  guides(
    shape = guide_legend(order = 1),
    colour = guide_legend(order = 2)
  ) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(vjust = 0.5),
        axis.title.x = element_text(vjust = -0.5),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        plot.title = element_text(size = 14, colour = "black", face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black", face = "bold"),
        legend.position = "inside",
        legend.position.inside = c(0.8, 0.64)) +
  labs(x = "Mean larval survival time / days (95% confidence interval)",
       y = "Species") 

# View figure
fig4b
# Save figure
ggsave("figures/fig4b.png", width = 12, height = 9)
```







